---
title: "Statistics Sweden"
author: "Göran Broström"
date: "February 2, 2018"
output: 
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::pdf_document2:
    citation_package: natbib
    number_sections: yes
    toc: no
    toc_depth: 2
    keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Studying ages 20--99, Sweden 1968--2014, for trend in expected life at 20 and median remaining life at 20. Using data from Statistics Sweden, stored in the **R** package **Sweden** .

# Get relevant data

```{r getdata}
library(Sweden)
fdeaths <- getDeaths(sex = "females", ages = 20:99)
fpop <- getPop(sex = "females", years = c(1968:2014), ages = 20:99)
mdeaths <- getDeaths(sex = "males", ages = 20:99)
mpop <- getPop(sex = "males", years = c(1968:2014), ages = 20:99)
```

Then we calculate yearly age-specific death rates:

```{r calcrates}
frates <- fdeaths / fpop
mrates <- mdeaths / mpop
```

And then we calculate $e_{20}$ and $m_{20}$ for the years 1968--2014.

# Statistics for men, age 20

```{r statmen, echo = FALSE}
library(eha)
source("R/mpch.R")
nYears <- NCOL(mrates)
me20 <- numeric(nYears)
mm20 <- numeric(nYears)
cts <- (21:99)
for (i in seq_len(nYears)){
    me20[i] <- mpch(cuts = cts - 20, levels = mrates[, i]) + 20
    mm20[i] <-  qpch(0.5, cuts = cts - 20, levels = mrates[, i]) + 20
}
plot(1968:2014, mm20, type = "b", main = "Men", ylim = c(70, 90), xlab = "Year", ylab = "Age", col = "red", las = 1)
lines(1968:2014, me20, col = "blue", type = "b")
text(1995, 82, "e.20", col = "red")
text(1995, 73, "m.20", col = "blue")
abline(h = c(75, 80, 85), lty = 3, col = "green")
abline(v = c(1970, 1980, 1990, 2000, 2010), lty = 3, col = "green")
mdiff <- mean(mm20 - me20)
```

We can also see that the difference between expected life and median life is fairly constant over time with a mean value of `r round(mdiff, 2)` years.

# Statistics for women, age 20

```{r statwomen, echo = FALSE}
library(eha)
source("R/mpch.R")
nYears <- NCOL(frates)
fe20 <- numeric(nYears)
fm20 <- numeric(nYears)
cts <- (21:99)
for (i in seq_len(nYears)){
    fe20[i] <- mpch(cuts = cts - 20, levels = frates[, i]) + 20
    fm20[i] <-  qpch(0.5, cuts = cts - 20, levels = frates[, i]) + 20
}
plot(1968:2014, fm20, type = "b", main = "Women", ylim = c(70, 90), xlab = "Year", ylab = "Age", col = "red", las = 1)
lines(1968:2014, fe20, col = "blue", type = "b")
text(1995, 86, "e.20", col = "red")
text(1995, 79, "m.20", col = "blue")
abline(h = c(75, 80, 85), lty = 3, col = "green")
abline(v = c(1970, 1980, 1990, 2000, 2010), lty = 3, col = "green")
fdiff <- mean(fm20 - fe20)
```

We can also see that the difference between expected life and median life is fairly constant over time with a mean value of `r round(fdiff, 2)` years.

