---
title: "fromFobLisa, Version 2"
author: "Göran Broström"
date: "March 1, 2018"
output:
  pdf_document: default
  html_document:
    df_print: paged
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
options(stringsAsFactors = FALSE)
library(stringr)
library(tidyr)
library(dplyr)
```

# Introduction

Now *fromFobLisa.Rmd* is updated to include *fob90* and also let the
follow-up for *fob85* be *five years* and the follow-up for *fob90* be *ten
years*.

This is an updated version of *showData4.Rmd*, where we used the newly-found
knowledge that the yearly *lisa* data relates to *December 31* the given
year, while the *FoB* data relate to the date *November 1* the given
year. These facts have the following implications for our plan: For *FoB*
data, we follow individuals from their appearance in a FoB until the next
FoB five years later, even if they do not appear there. The exception is if
they die within five years after the first appearance. So, *fob60* induces
a time interval *from November 1, 1960 to October 31, 1965*, and so on. 
Since we do not have birth dates but only birth year and quarter,
we decide to use as birthdate the *midpoint of the birth quarter*. 
WE work with all ages 0-110, and only as the final step, `age.window` is called.

The following FoBs are treated similarly, until we reach the Fob
in 1985. Since the first *lisa* date is December 31, 1986, this "fob
interval" starts at November 1, 1985, and ends at December 31, 1986, that
is, it is 14 months long. 

# Fixed data

We start with the *inddata* file and treat it as a *fixed-data*
file (no time-varying info). 
We find the birth dates (see above), death dates (for those who are observed die), and the "date last seen".

Note that 37 individuals in *inddata* have no information at all, so they are removed.

```{r frominddata}
load("../../Data/inddata.rda")
load("../../Data/sex.rda")
indx <- match(inddata$LINNEID, sex$LINNEID)
inddata$KON <- sex$KON[indx]
per <- inddata[!is.na(inddata$KON) & !is.na(inddata$FODELSEAR), ]
per$birthdate <- per$FODELSEAR + 0.125  +  (per$FODELSEKVARTAL - 1) / 4
```

Now we put on *death date*, *death cause*, and *KOMMUN of death*, from *socstyr*, on *per*:

```{r styronper}
load("../../Data/socstyr.rda")
indx <- match(per$LINNEID, socstyr$LINNEID)
per$DODSDAT <- socstyr$DODSDAT[indx]
fixd <- !is.na(per$DODSDAT) & (per$DODSDAT < 1960010)
per$DODSDAT[fixd] <- per$DODSDAT[fixd] * 100 + 15
per$deathdate <- as.numeric(skum::tillTid(per$DODSDAT))
per$dodsKOMMUN <- socstyr$KOMMUN[indx]
per$dodsALDER <- socstyr$ALDER[indx]
per$ULORSAK <- socstyr$ULORSAK[indx]
```

# The FoBs

Through the *FoB*s we have snapshots of the population on *November 1* the years *1960(5)1990*, giving the dates (decimal form).

```{r days}
load("../../Data/fob.rda")
fob <- fob[fob$LINNEID %in% per$LINNEID, ]
ar <- seq(1960, 1990, by = 5)
days <- as.numeric(skum::tillTid(paste(ar, "1101", sep = "")))
days
```

We have to limit the data collection to the *KOMMUN*er that are contained
in our area, that is

```{r kommuner}
ourkom <- c(2406, 2414, 2417, 2418, 2480, 2482)
ourkom <- c(ourkom, 2411, 2412, 2413, 2416, 2415, 2461, 2407, 2402, 2408)
fob <- fob[fob$KOMMUN %in% ourkom, ]
```

Get `birthdate` from `per`.

```{r getbirthdateanddathdate}
indx <- match(fob$LINNEID, per$LINNEID)
fob$birthdate <- per$birthdate[indx]
fob$deathdate <- per$deathdate[indx]
```

Create individual Fobs (1960, 1965, etc):

```{r rawfob}
source("R/extractFob.R")
n <- length(ar)
fobbar <- vector(mode = "list", length = n)
names(fobbar) <- paste("fob", (ar %% 100), sep = "")
for (i in 1:(n-1)){
    fobbar[[i]] <- extractFob(ar[i], followUp = 5, data = fob)
}
fobbar[[n]] <- extractFob(ar[n], followUp = 11 + 61 / 365, data = fob)
fobdata <- do.call(rbind, fobbar)
fobdata <- fobdata[order(fobdata$LINNEID, fobdata$enter), ]
##
rownames(fobdata) <- 1:NROW(fobdata)
```

# Fob gets hisclass and the rest

We continue as in "Mort2/CreateData2/fromFobLisa.Rmd":

```{r syskonfob}
    load("../../Data/socstyr.rda")
    indx <- match(fobdata$LINNEID, socstyr$LINNEID)
    fobdata$ULORSAK <- str_trim(as.character(socstyr$ULORSAK[indx]))
    source("R/putONssyk.R")
    ##putONssyk
    fobb <- putONssyk(fobdata)
    source("R/ssykTOhisclass.R")
    ##ssykTOhisclass
    fobb$hisclass <- ssykTOhisclass(fobb$ssyk)
    is.na(fobb$hisclass) <- with(fobb, is.na(hisclass) |
                                       (!is.na(hisclass) & hisclass == -1))
    fobb$id <- fobb$LINNEID
    load("../../Data/sex.rda")

    indx <- match(fobb$LINNEID, sex$LINNEID)
    fobb$KON <- sex$KON[indx]
    fobb$urban <- fobb$FORSAMLING %in% c(248001, 248002, 284004, 284005,
                                         248011, 248201, 248204)
    civst <- numeric(NROW(fobb))
    civst[fobb$CIVIL %in% c(0, 5)] <- 1 # Ogift
    civst[fobb$CIVIL %in% c(2, 7)] <- 2 # Gift, sambo
    civst[fobb$CIVIL %in% c(1, 3, 4, 6, 8, 9)] <- 3 # skild, änka
    fobb$civst <- factor(civst,
                         labels = c("unmarried", "married", "prev.married"))
    fobb$sex <- fobb$KON
    levels(fobb$sex) <- c("male", "female")
    fobb <- fobb[, c("id", "AR", "birthdate", "sex", "enter", "exit", "event",
                     "urban", "civst", "hisclass", "ULORSAK")]
```

# The Lisa data

The *lisa* data is slightly simpler because it is yearly with cutoff dates 12-31/01-01 each new-year. It starts off at 1987-01-01 (lisa1986) and ends at 2013-12-31 (lisa2013). So the time interval a *lisa* covers is *AR + 1*! But note that the *socstyr* data only cover deaths *before January 1, 2014, so *lisa2013* is in effect useless to us.  

```{r lisadat}
load("../../Data/lisa.rda")
lisa <- lisa[(lisa$KOMMUN %in% ourkom) & (lisa$LINNEID %in% per$LINNEID), ]
indx <- match(lisa$LINNEID, per$LINNEID)

lisa$birthdate <- per$birthdate[indx]

```

So, in *lisa*, all enter in *AR + 1*, at the exact age of *(AR + 1 - birthdate)*, and exit at the exact age *enter + 1* if they do not die in *the year AR + 1*, otherwise at their death date, or at *deathdate - birthdate*.

```{r fixlisaintervals}
lisa$enter <- lisa$AR + 1 - lisa$birthdate
indx <- match(lisa$LINNEID, per$LINNEID) # To be safe ...
lisa$deathdate <- per$deathdate[indx]
lisa$event <- !is.na(lisa$deathdate) &
    lisa$deathdate > lisa$AR + 1 & 
    lisa$deathdate <= lisa$AR + 2
lisa$exit <- lisa$enter + 1 # First tip.
lisa$exit[lisa$event] <- with(lisa, deathdate[event] - birthdate[event])
```

Now some sorting  and cleaning.

```{r cleanandsort}
lisadata <- lisa[order(lisa$LINNEID, lisa$enter), ]
lisadata <- cal.window(lisadata, c(1987, 2014)) # 'Remove lisa2013'
rownames(lisadata) <- 1:NROW(lisadata)
```

## Lisa gets hisclass and the rest


```{r lisadata}
    ## Cause of death:
    load("../../Data/socstyr.rda")
    indx <- match(lisadata$LINNEID, socstyr$LINNEID)
    lisadata$ULORSAK <- str_trim(as.character(socstyr$ULORSAK[indx]))
    ## Civil status:
    civst <- lisadata$CIV
    civst[is.na(civst)] <- 0
    civst[civst == 4] <- 3
    civst[civst == 5] <- 2
    civst[civst == 6] <- 3
    is.na(civst) <- civst == 0
    lisadata$civst <- civst
    rm(civst)
    ## Fix KON:
    load("../../Data/sex.rda")
    indx <- match(lisadata$LINNEID, sex$LINNEID)
    lisadata$sex <- sex$KON[indx]
    levels(lisadata$sex) <- c("male", "female")
    lisaold <- lisadata
    lisadata <- cal.window(lisadata, c(2002, 2014))

    ## Fix HISCLASS:
    source("R/ssyk3TOhisclass.R")
    ##ssyk3TOhisclass
    lisadata$hisclass <- ssyk3TOhisclass(lisadata$SSYK3)
    liss <- lisadata

    liss$id <- liss$LINNEID
    liss$civst <- factor(liss$civst,
                         labels = c("unmarried", "married", "prev.married"))
 
    liss$urban <- liss$FORSAMLING %in% c(248001, 248002, 284004,
                                         284005, 248011, 248201, 248204)
 
    liss$id <- liss$LINNEID
    liss <- liss[, c("id", "AR", "birthdate", "sex", "enter", "exit", "event",
                     "urban", "civst", "hisclass", "ULORSAK")]
 
```


# Combine Fob and Lisa


```{r combfoblisa}

foli <- rbind(fobb, liss)
foli <- foli[order(foli$id, foli$enter), ]
source("R/pushForward.R")
foli$hisclass <- unlist(tapply(foli$hisclass, foli$id, pushForward))
foli$hisclass[is.na(foli$hisclass)] <- 5
foli$hisclass <- factor(foli$hisclass, 
                        labels = c("elite", "middle", "worker", "farmer", "none"))
zz <- levels(foli$civst)
foli$civst <- unlist(tapply(as.integer(foli$civst), foli$id, pushForward))
foli$civst <- factor(foli$civst, labels = zz)

foli <- age.window(foli, c(30, 100))
save(foli, file = "../data/foli.rda")
```

